### 2020-12-10 submission/manuscript.tex
Final edits

\hypertarget{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}{%
\section{Removal of rare sequences from 16S rRNA gene sequence surveys
biases the interpretation of community structure
data}\label{removal-of-rare-sequences-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}}

@@

\textbf{word choice: sample vs.~community} \textbf{word choice: dataset
vs.~sample}

@@

Methods for remediating PCR and sequencing artifacts in 16S rRNA gene
sequence collections are in continuous development. The choices that
researchers make have significant ramifications on the inferences that
can be drawn from microbial ecology studies. One common approach is to
remove rare sequences from datasets. But, the threshold that is applied
is generally selected without regard for the number of sequences in the
sample or the variation in sequencing depth across samples within a
study. I analyzed the impact of removing rare sequences on metrics of
alpha and beta diversity using samples collected across 12 published
datasets. Removal of rare sequences significantly decreased the number
of amplicon sequence variants and operational taxonomic units as well as
their diversity. Furthermore, their removal increased the variation in
community structure between samples. When a known effect size was
simulated, removal of rare sequences reduced the statistical power to
detect the effect relative to not removing rare sequences. Although
removal of rare sequences did not affect the probability of falsely
detetecting differences when samples were randomized to simulate a null
model, the false detection rate increased when samples were assigned to
simulated treatment groups according to their sequencing depth and rare
sequences were removed. The false detection rate did not vary when rare
sequences were retained. This analysis demonstrates the problems
inherent in removing rare sequences. Researchers are encouraged to
retain these sequences, to select approaches that minimize PCR and
sequencing artifacts, and to use rarefaction to control for uneven
sampling effort.

@@

Removing rare sequences from 16S rRNA gene sequence collections is an
approach that has grown in popularity for handling PCR and sequencing
artifacts. Yet, it is unclear what impact an abundance-based filter has
on downstream analyses. To investigate the effects of removing rare
sequences, I analyzed the community distributions found in the samples
of 12 published datasets. Analysis of these data and simulations based
on them showed that removal of rare sequences distorts the
representation of microbial communities. This has the effect of
artificially making it more difficult to detect differences between
treatment groups. Also of concern was the observation that if sequencing
depth is confounded with treatment group, then the probability of
falsely detecting a difference between the treatment groups increases
with the removal of rare sequences. The practice of removing rare
sequences should stop, lest researcher adversely affect the
interpretation of their data.

@@

Numerous studies have attempted to address the problem of removing or
``denoising'' artifacts from data generated using Illumina's MiSeq
platform. In one approach, paired sequence reads are aligned and any
discrepencies between the reads are resolved based on the difference in
quality score for the position in question (2--4); quality scores are
also used to curate single reads (5). In addition, a polishing step is
often used to identify ideal sequecnces and their related artifacts
based on the frequency of the sequences and their similarity sequences
(2, 3, 6). In a second approach, the quality scores and types of errors
to cluster sequences directly (7). Regardless of the approach, many
pipelines advocate for abundance-based sceeening where rare sequences
are removed from each dataset prior to outputting the sequence data as
ASVs (3, 6, 7). Some algorithms recommend removing all ASVs that appear
one (i.e.~singletons) (7), eight (3), or ten (6) or fewer times; these
pipelines also vary in whether the minimum abundance threshold should be
applied to individual samples (6, 7) or the pool of samples in a study
(3). A notable exception, the mothur-based pipeline discourages the
practice of removing rare sequences (2). After assigning sequences to
ASVs, they are often analyzed as a taxonomic unit or clustered to
generate operational taxonomic units (OTUs) or phylotypes.

@@

The minimum abundance thresholds that have been proscribed were
developed and applied without regard for the number of sequences
generated from each sample. These recommendations appear to assume that
the number of reads per sample is consistent within and between studies.
However, it is common that the number of sequences generated from each
sample may vary by two or three orders of magnitude (e.g.~Figure S1). An
ASV that appears once in a sample with 2,000 sequences is more
trustworthy than an ASV that appears once in a sample with 100,000
sequences since it has a 50-fold higher relative abundance. But,
according to the pipeline recommendations, they are treated as being
equally trustworthy. Rather than removing rare ASVs, the approach taken
by the mothur pipeline applies the classical ecological approach of
rarefaction. Each sample is rarefied to the same sequencing depth so
that the number of artifacts that appears in each sample is controlled.

Experience sequencing biological samples also demonstrates that there
are bona fide ASVs that may have an abundance below the proscribed
threshold. For example, the abundance of an ASV may be below the
threshold in some samples or time points and above the threshold in
others. However, rarity, both in terms of prevalence and incidence, is
an important ecological concept. Removing rare ASVs likely hinders one's
to ability to make inferences about the dynamics and nature of the
populations that rare ASVs represent. Furthermore, removing ASVs whose
abundances are below the proscribed threshold also potentially biases
the community structure of the samples.

@@

\textbf{Datasets.} I collected 12 publicly available datasets that used
the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA
gene from a variety of environments (Table 1). After removing poor
quality and chimeric ASVs and samples that had uncharacteristically low
number of sequences for the dataset, these datasets included between 7
and 490 samples (Figure S1). The median number of sequences for each
dataset ranged between 6,477 and 193,464 (Table 1). Strikingly, aside
from the relatively small marine and soil datasets, the difference
between the sample with the fewest sequences and the sample with the
most sequences for each dataset varied by between 7.4 and 96.6-fold
(Table 1).

@@

\textbf{Removing treatment group effects from community data.} Because
treatment effects often affect a sample's diversity and inter-sample
variation, I generated null distributions for each study by randomizing,
without replacement, the number of times each ASV was observed in each
sample such that the total number of sequences in each sample and the
total number of times each ASV was observed across all samples in the
study was the same as was originally observed. This effectively made
every community in a study a statistical sample of the study-wide
composite community distribution. For example, after this procedure, the
490 samples from the human dataset would be expected to have the same
richness and diversity of ASVs and one would not expect to find
treatment-based effects between the samples. Because of the risk of bias
if only one representation of the null distribution was generated, I
generated 100 randomized datasets for each study. The trends between
removing rare ASVs and the richness, diversity, and information loss
that were identified using the observed community community distribution
data were also identified with the data from the null distribution;
however, the losses were larger when using the null distribution data
(Figure S3). The null distribution data were used in the remainder of
the study to minimize the risk of bias.

@@

\textbf{The impact of removing rare ASVs on the ability to detect
statistically significant differences between treatment groups.} To test
the effect of increased inter-sample variation, I randomly assigned
samples to one of two treatment groups. In the first treatment group,
communities were randomly sampled from the null distribution as
described above. For the second treatment group, I randomly selected
10\% of the ASVs in the pooled study distribution to increase their
abundance by 5\%. I randomly generated 100 simulated sets of treatment
groups and samples. I then tested the ability to detect a difference
between the two treatment groups using alpha and beta diversity metrics.
The fraction of significant tests was a measurement of the statistical
power to detect the difference between the treatment groups. When
considering the differences in richness and diversity, the marine
dataset yielded no simulated sets that were statistically significant,
which was likely due to the small number of samples in the study (N=7).
Among the remaining datasets, the power to detect a difference in the
richness of ASVs ranged between 0.10 and 0.49 (sediment and stream) and
between 0.10 and 0.53 (rainforest and stream) to detect a difference in
diversity when using a Wilcox test (Figure 4A). When singleton ASVs were
removed, the power to detect a difference in the diversity of ASVs
dropped by between 27.3 and 92.9\% (bioethanol and soil) and by between
40.0 and 93.3\% (rainforest and soil; Figure 4B). The effect of removing
rare ASVs on the richness of OTUs and their diversity was similar
(Figure S5AB). I used the Bray-Curtis dissimilarity index to compare the
simulated communities within each dataset and calculated the power to
detect differences between the two simulated treatment groups using the
analysis of molecular variance (Figure 4C and S5C). Without removing
rare sequences, the power to detect a difference between the two
simulated treatment groups varied between 0.41 and 1.00 (rainforest and
rainforest). Aside from the bioethanol, human, and mice datasets, the
power to detect differences dropped by between 6.5 and 64.0\% (soil and
rice) when singletons were removed. However, when ASVs that occurred 10
or fewer times were removed from each sample, the power to detect
differences dropped by 12.0 and 97.2\% (human and peromyscus); similar
results were observed when ASVs were clustered into OTUs. Removing rare
ASVs reduced the ability to detect simulated treatment effects using
metrics commonly used to compare microbial communities.

\textbf{The impact of removing rare ASVs on the probability of falsely
detecting a difference between treatment groups.} I next asked whether
removing rare ASVs could lead to falsely claiming that a treatment
effect had a significant effect on community diversity and structure.
First, I sampled sequences from the null distribution for each dataset
and randomly assigned each sample to one of two treatment groups and
determined the richness and diversity of ASVs and OTUs. Testing at an
experiment-wise error rate of 0.05, I expected 5\% of the iterations for
each dataset to yield a significant test result. Indeed, there was no
evidence that removing rare ASVs resulted in an inflated experiment-wise
error rate. The average fraction of significant tests did not
meaningfully vary from 0.05 across the minimum abundance threshold,
dataset, metric of describing sample alpha-diversity, or whether the
abundance of ASVs or OTUs were used (Figure 5A and S6A). Similarly, the
average fraction of significant tests did not meaningfully vary from
0.05 when using analysis of molecular variance to compare communities
using Bray-Curtis distances (Figure 5A and S6A). Second, I again sampled
sequences from the null distribution, but assigned samples to one of two
treatment groups based on the number of sequences in each sample. The
samples with fewer than the median number of sequences for the dataset
were assigned to one group and those with more than the median were
assigned to the other. This exaggerated bias has been observed in
comparisons of the lung and oral microbiota because of the larger number
of non-specific amplicons that can be sequenced from lung samples
relative to those in the oral cavity leading to a significant difference
in sequencing depth between treatment groups (8). When rare sequences
were not removed, the fraction of significant tests did not differ from
5\% for comparing the richness, their diversity, or Bray-Curtis
distances (Figure 5B and S6B). However, when rare taxa of any frequency
were removed, the probability of falsely detecing a difference as
signifiant increased with the definition of rarity (Figure 5B and S6B).
Not including the small marine dataset, the average fraction the average
fraction of falsely detecting a difference across datasets when only
singletons were removed was 92.45\%. If there is any relationship
between the number of sequences and the treatment group, the risk of
falsely rejecting the null hypothesis is inflated when researchers use
the strategy of removing rare sequences. The most conservative approach
is to not remove low abundance sequences.

@@

The practice of removing rare sequences from samples seems to be a
response to researchers prioritizing the number of reads and length of
sequences over their quality (5). Previous work has shown that assembly
of fully overlapping sequence reads results in the lowest sequencing
error rates (2). The studies highlighted in this analysis sequenced the
V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing
platform with their v2 chemistry. The resulting data consists of two 250
nt reads that span a region that is about 250 nt long. In contrast,
there has been a movement to sequence longer regions with similar
chemistry resulting in less overlap between the sequencing reads (9,
10). Alternatively, others have prioritized increasing the number of
sequences per sample by sequencing the V4 region, but with paired or
single reads that are 150 nt long (11, 12). Both practices result in a
significantly higher error rate for the resulting assmebly. Instead,
researchers should prioritize the quality over the quantity and length
of their data. For these reasons, this analysis did not use lower
quality data generated by alternative methods

The impacts of removing rare sequences on the representation of
communities are caused by the uneven impact of applying a single
abundance threshold across all samples. The number of sequences per
sample in the datasets highlighted in this analysis varied by 100-fold
(Figure S1); I suspect the range is actually larger since researchers
may have opted against depositing samples with fewer reads into the SRA.
If the range in read coverage for a study is 100-fold, a singleton in a
small sample would have had a comparable relative abundance as a
sequence that appeared 100 times in a more densely sequenced sample.
However, it would have been removed from the smaller sample and not the
larger sample. Thus, applying a single abundance threshold
disproportionately impacted the samples with fewer sequences. This seems
to contradict the purpose of removing rare sequences since a singleton
in the smaller sample is more reliable than the singleton in the larger
sample. A superior approach to removing rare sequences is to use
rarefaction to conrol for uneven sampling.

Even with the best sequencing approaches, PCR artifacts and sequencing
errors persist. This may account for some sequences not being observed
other samples. Although this lack of inter-sample coverage could have
been due to treatment effects and natural variation, it is likely that
some portion of the sequences that were unique to samples were artifacts
or contained errors. Researchers are encouraged to use rarefaction to
control for uneven sampling and the presence of spurious sequences.
Previous work sequencing mock communities has shown that the number of
spurious sequences increases with sequencing depth (2, 13). By rarefying
data to a common number of sequences per sample, the number of spurious
sequences can be controlled. As shown in the data I presented, which was
rarefied to a common number of reads per sample within a dataset, when
rare sequences were not removed the power to detect differences was the
highest and the false discovery rate was the expected 5\% (Figures 5 and
S6).

In addition to considerations of how to control for the presence of
spurious sequences, researchers also need to be mindful of how to
interpret the results of their work. Because every dataset will contain
residual sequences that are spurious, measures of richness and diversity
should be made on a relative basis. For example, pronouncements that
communities from an environment or treatment contain a specific number
of taxa are problematic. Instead, we should limit ourselves to
indicating that samples from one treatment group has more taxa than
another without using absolute values of richness. Furthermore, we must
take caution in interpreting rare taxa. Although the data from the
studies highlighted here suggest that most rare sequences are not
spurious, it is likely that some are. Therefore, researchers must
approach rare sequences with more skepticism than more abundant
sequences. Reseachers should seek out other methods to confirm
inferences that they make about rare sequences. This is a standard that
should be applied regardless of their abundance (14).

How to curate and interpret rare sequences has been a significant
challenge since microbial ecologists transitioned away from Sanger
sequencing of samples (15--17). Although the extent of the ``rare
biosphere'' is still an open question, it is important to appreciate the
importance of rare populations in all communities. Populations can be
numerically rare but ubuiquitous or abundant and limited in their
geographic range. Alternatively, they can be numerically rare but
temporally common or abundant but present infrequently. Removing
sequences from any of these settings will limit our ability to study the
role of such populations or the processes that drive their patchy
distributions that are so common with microbial communities (18).

@@

\textbf{Data curation and analysis.} To insure the highest possible data
quality, datasets were limited to those where the 500 cycle v2 MiSeq
chemistry was used to sequence the amplicons. The paired 250 nt reads
resulted in near complete 2-fold sequencing coverage of every nucleotide
in the ca. 250 nt-long region. This region and sequencing platform were
selected because previous work has shown that a standard data analysis
pipeline in mothur results in a sequencing error rate below 0.02\% (2).
All sequence data were obtained from the Sequence Read Archive (SRA) and
processed using a standard mothur-based sequencing pipeline that
resulted in ASVs as generated by the pre.cluster algorithm using a
threshold of 2 nt (2, 19). ASVs were assigned to OTUs using a 3\%
distance threshold using mothur's cluster function with the OptiClust
algorithm (20). To minimize the effects of uneven sampling effort,
samples were rarefied to the number of sequences in the smallest sample
for each dataset. Because metrics of alpha diversity did not
consistently follow a normal distribution, I used the non-parametric
Wilcoxon rank test as implemented in R. For comparisons of Bray-Curtis
distances, the amova function within mothur was used, which implements
the analysis of molecular variance algorithm (21).

\textbf{Reproducibility.} All analyses were performed using mothur
(version 1.44.1) and R (version 4.0.2) with the tidyverse (v.130), broom
(v.070), data.table (v. 113 4), and cowplot (v.110) packages. All of the
code that was used in this analysis as well as the Makefile for running
the analysis are available at
\url{https://github.com/SchlossLab/Schloss_Singletons_XXXXX_2019} along
with the complete version history of the project.

@@

\leavevmode\hypertarget{ref-Morris2013}{}%
\CSLLeftMargin{8. }
\CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
PD}, \textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
\textbf{Flores SC}, \textbf{Fontenot AP}, \textbf{Ghedin E},
\textbf{Huang L}, \textbf{Jablonski K}, \textbf{Kleerup E},
\textbf{Lynch SV}, \textbf{Sodergren E}, \textbf{Twigg H}, \textbf{Young
VB}, \textbf{Bassis CM}, \textbf{Venkataraman A}, \textbf{Schmidt TM},
\textbf{Weinstock GM}. 2013. Comparison of the respiratory microbiome in
healthy nonsmokers and smokers. American Journal of Respiratory and
Critical Care Medicine \textbf{187}:1067--1075.
doi:\href{https://doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.}

\leavevmode\hypertarget{ref-Fadrosh2014}{}%
\CSLLeftMargin{9. }
\CSLRightInline{\textbf{Fadrosh DW}, \textbf{Ma B}, \textbf{Gajer P},
\textbf{Sengamalay N}, \textbf{Ott S}, \textbf{Brotman RM},
\textbf{Ravel J}. 2014. An improved dual-indexing approach for
multiplexed 16S {rRNA} gene sequencing on the illumina {MiSeq} platform.
Microbiome \textbf{2}:6.
doi:\href{https://doi.org/10.1186/2049-2618-2-6}{10.1186/2049-2618-2-6}.}

>>>>

\hypertarget{removal-of-rare-amplicon-sequence-variants-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}{%
\section{Removal of rare amplicon sequence variants from 16S rRNA gene
sequence surveys biases the interpretation of community structure
data}\label{removal-of-rare-amplicon-sequence-variants-from-16s-rrna-gene-sequence-surveys-biases-the-interpretation-of-community-structure-data}}

@@

@@

Methods for remediating PCR and sequencing artifacts in 16S rRNA gene
sequence collections are in continuous development and have significant
ramifications on the inferences that can be drawn. A common approach is
to remove rare amplcon sequence variants (ASVs) from datasets. But, the
definition of rarity is generally selected without regard for the number
of sequences in the samples or the variation in sequencing depth across
samples within a study. I analyzed the impact of removing rare ASVs on
metrics of alpha and beta diversity using samples collected across 12
published datasets. Removal of rare ASVs significantly decreased the
number of ASVs and operational taxonomic units as well as their
diversity. Furthermore, their removal increased the variation in
community structure between samples. When simulating a known effect
size, removal of rare ASVs reduced the power to detect the effect
relative to not removing rare ASVs. Removal of rare ASVs did not affect
the false detection rate when samples were randomized to simulate a null
model. However, the false detection rate increased when rare ASVs were
removed using a null distribution and assignment of samples to simulated
treatment groups according to their sequencing depth. The false
detection rate did not vary when rare ASVs were retained. This analysis
demonstrates the problems inherent in removing rare ASVs. Researchers
are encouraged to retain rare ASVs, to select approaches that minimize
PCR and sequencing artifacts, and to use rarefaction to control for
uneven sequencing effort.

@@

Removing rare amplicon sequence variants (ASVs) from 16S rRNA gene
sequence collections is an approach that has grown in popularity for
limiting PCR and sequencing artifacts. Yet, it is unclear what impact an
abundance-based filter has on downstream analyses. To investigate the
effects of removing rare ASVs, I analyzed the community distributions
found in the samples of 12 published datasets. Analysis of these data
and simulations based on them showed that removal of rare ASVs distorts
the representation of microbial communities. This has the effect of
artificially making it more difficult to detect differences between
treatment groups. Also of concern was the observation that if sequencing
depth is confounded with the treatment, then the probability of falsely
detecting a difference between the treatment groups increased with the
removal of rare ASVs. The practice of removing rare ASVs should stop,
lest researcher adversely affect the interpretation of their data.

@@

Numerous studies have attempted to address the problem of removing or
``denoising'' artifacts from data generated using Illumina's MiSeq
platform. In one approach, paired sequence reads are aligned and any
discrepencies between the reads are resolved based on the difference in
quality score for the position in question (2--4); quality scores are
also used to curate single reads (5). In addition, a polishing step is
often used to identify ASVs based on the frequency and similarity of
sequences (2, 3, 6). In a second approach, the quality scores and types
of errors are modelled to cluster sequence reads directly into ASVs (7).
Regardless of the approach, many pipelines advocate for abundance-based
screening where rare sequences are removed from each dataset prior to
outputting the sequence data as ASVs (3, 6, 7). Some algorithms
recommend removing all ASVs that appear one (i.e.~singletons) (7), eight
(3), or ten (6) or fewer times; these pipelines also vary in whether the
minimum abundance threshold should be applied to individual samples (6,
7) or the pool of samples in a study (3). A notable exception, the
mothur-based pipeline discourages the practice of removing rare
sequences (2). After assigning sequences to ASVs, ASVs are often
analyzed as a taxonomic unit or clustered to generate operational
taxonomic units (OTUs) or phylotypes.

@@

The minimum abundance thresholds that have been proscribed were
developed and applied without regard for the number of sequences
generated from each sample. These recommendations appear to assume that
the number of reads per sample is consistent within and between studies.
However, it is common that the number of sequences generated from each
sample may vary by two or three orders of magnitude (e.g.~Table 1 and
Figure S1). An ASV that appears once in a sample with 2,000 sequences is
more trustworthy than an ASV that appears once in a sample with 100,000
sequences since it has a 50-fold higher relative abundance. But,
according to the pipeline recommendations, they are treated as being
equally trustworthy. Rather than removing rare ASVs, the approach taken
by the mothur pipeline applies the classical ecological approach of
rarefaction. Each sample is rarefied to the same sequencing depth so
that the number of artifacts that appears in each sample is controlled.

Experience sequencing biological samples also demonstrates that there
are bona fide ASVs that may have an abundance below the proscribed
threshold. For example, the abundance of an ASV may be below the
threshold in some samples or time points and above the threshold in
others. However, rarity, both in terms of prevalence and incidence, is
an important ecological concept (8). Removing rare ASVs likely hinders
one's to ability to make inferences about the dynamics and nature of the
populations that rare ASVs represent. Furthermore, removing ASVs whose
abundances are below the proscribed threshold also potentially biases
the community structure of the samples.

@@

\textbf{Datasets.} I collected 12 publicly available datasets that used
the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA
gene from a variety of environments (Table 1; Figure S1). After removing
poor quality and chimeric ASVs and samples that had uncharacteristically
low number of sequences for the dataset, these datasets included between
7 and 490 samples. The median number of sequences for each dataset
ranged between 6,477 and 193,464. Strikingly, aside from the relatively
small marine and soil datasets, the difference between the sample with
the fewest sequences and the sample with the most sequences for each
dataset varied by between 7.4 and 96.6-fold.

@@

\textbf{Removing treatment group effects from community data.} Because
treatment effects often affect a sample's diversity and inter-sample
variation, I generated null distributions for each study by randomizing,
without replacement, the number of times each ASV was observed in each
sample such that the total number of sequences in each sample and the
total number of times each ASV was observed across all samples in the
study was the same as was originally observed. This effectively made
every community in a study a statistical sample of the study-wide
composite community distribution. For example, after this procedure,
each of the 490 samples from the human dataset would be expected to have
the same richness and diversity of ASVs and one would not expect to find
treatment-based effects between the samples. Because of the risk of bias
if only one representation of the null distribution was generated, I
generated 100 randomized datasets for each study. The trends between
removing rare ASVs and the richness, diversity, and information loss
that were identified using the observed community distribution data were
also identified with the data from the null distribution; however, the
losses were larger when using the null distribution data (Figure S3).
The null distribution data were used in the remainder of the study to
minimize the risk of bias.

@@

\textbf{The impact of removing rare ASVs on the ability to detect
statistically significant differences between treatment groups.} To test
the effect of increased inter-sample variation, I randomly assigned
samples to one of two treatment groups. In the first treatment group,
communities were randomly sampled from the null distribution as
described above. For the second treatment group, I randomly selected
10\% of the ASVs in the pooled study distribution to increase their
abundance by 5\%. I randomly generated 100 simulated sets of treatment
groups and samples. I then tested the ability to detect a difference
between the two treatment groups using alpha and beta diversity metrics.
The fraction of significant tests was a measurement of the statistical
power to detect the difference between the treatment groups. When
considering the differences in richness and diversity, the marine
dataset yielded no simulated sets that were statistically significant,
which was likely due to the small number of samples in the study (N=7).
Among the remaining datasets, the power to detect a difference in the
richness of ASVs ranged between 0.10 and 0.49 (sediment and stream) and
between 0.10 and 0.53 (rainforest and stream) to detect a difference in
diversity when using a Wilcox test (Figure 4A). When singleton ASVs were
removed, the power to detect a difference in the richness of ASVs
dropped by between 27.3 and 92.9\% (bioethanol and soil) and by between
40.0 and 93.3\% for their diversity(rainforest and soil; Figure 4B). The
effect of removing rare ASVs on the richness of OTUs and their diversity
was similar (Figure S5AB). I used the Bray-Curtis dissimilarity index to
compare the simulated communities within each dataset and calculated the
power to detect differences between the two simulated treatment groups
using the analysis of molecular variance (Figure 4C and S5C). Without
removing rare sequences, the power to detect a difference between the
two simulated treatment groups varied between 0.41 and 1.00 (rainforest
and rainforest). Aside from the bioethanol, human, and mice datasets,
the power to detect differences dropped by between 6.5 and 64.0\% (soil
and rice) when singletons were removed. However, when ASVs that occurred
10 or fewer times were removed from each sample, the power to detect
differences dropped by 12.0 and 97.2\% (human and peromyscus); similar
results were observed when ASVs were clustered into OTUs. Removing rare
ASVs reduced the ability to detect simulated treatment effects using
metrics commonly used to compare microbial communities.

\textbf{The impact of removing rare ASVs on the probability of falsely
detecting a difference between treatment groups.} I next asked whether
removing rare ASVs could lead to falsely claiming that a treatment
effect had a significant effect on community diversity and structure.
First, I sampled sequences from the null distribution for each dataset
and randomly assigned each sample to one of two treatment groups and
determined the richness and diversity of ASVs and OTUs. Testing at an
experiment-wise error rate of 0.05, I expected 5\% of the iterations for
each dataset to yield a significant test result. Indeed, there was no
evidence that removing rare ASVs resulted in an inflated experiment-wise
error rate. The average fraction of significant tests did not
meaningfully vary from 0.05 across the minimum abundance threshold,
dataset, metric of describing sample alpha-diversity, or whether the
abundance of ASVs or OTUs were used (Figure 5A and S6A). Similarly, the
average fraction of significant tests did not meaningfully vary from
0.05 when using analysis of molecular variance to compare communities
using Bray-Curtis distances (Figure 5A and S6A). Second, I again sampled
sequences from the null distribution, but assigned samples to one of two
treatment groups based on the number of sequences in each sample. The
samples with fewer than the median number of sequences for the dataset
were assigned to one group and those with more than the median were
assigned to the other. This exaggerated bias has been observed in
comparisons of the lung and oral microbiota because of the larger number
of non-specific amplicons that can be sequenced from lung samples
relative to those in the oral cavity leading to a significant difference
in sequencing depth between treatment groups (9). When rare sequences
were not removed, the fraction of significant tests did not differ from
5\% for comparing the richness, their diversity, or Bray-Curtis
distances (Figure 5B and S6B). However, when rare taxa of any frequency
were removed, the probability of falsely detecing a difference as
signifiant increased with the definition of rarity (Figure 5B and S6B).
Not including the small marine dataset, the average fraction the average
fraction of falsely detecting a difference across datasets when only
singletons were removed was 92.45\%. If there is any relationship
between the number of sequences and the treatment group, the risk of
falsely rejecting the null hypothesis is inflated when researchers use
the strategy of removing rare sequences. The most conservative approach
is to not remove low abundance sequences.

@@

The practice of removing rare sequences from samples seems to be a
response to researchers prioritizing the number of reads and length of
sequences over their quality (5). Previous work has shown that assembly
of fully overlapping sequence reads results in the lowest sequencing
error rates (2). The studies highlighted in this analysis sequenced the
V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing
platform with their version 2 chemistry. The resulting data consists of
two 250 nt reads that span a region that is about 253 nt long. In
contrast, there has been a movement to sequence longer regions with
similar chemistry resulting in less overlap between the sequencing reads
(10, 11). Alternatively, others have prioritized increasing the number
of sequences per sample by sequencing the V4 region, but with paired or
single reads that are 150 nt long (12, 13). Both practices result in a
significantly higher error rate for the resulting assmebly. Instead,
researchers should prioritize the quality over the quantity and length
of their data. For these reasons, this analysis did not use lower
quality data generated by alternative methods.

The impacts of removing rare sequences on the representation of
communities are caused by the uneven impact of applying a single
abundance threshold across all samples. The number of sequences per
sample in the datasets highlighted in this analysis varied by as much as
100-fold (Table 1; Figure S1). In practice, I suspect this range is
actually larger since researchers may have opted against depositing
samples with fewer reads into the SRA. If the range in read coverage for
a study is 100-fold, a singleton in a small sample would have had a
comparable relative abundance as a sequence that appeared 100 times in a
more densely sequenced sample. However, it would have been removed from
the smaller sample and not the larger sample. Thus, applying a single
abundance threshold disproportionately impacted the samples with fewer
sequences. This seems to contradict the purpose of removing rare
sequences since a singleton in the smaller sample is more reliable than
the singleton in the larger sample. A superior approach to removing rare
sequences is to use rarefaction to conrol for uneven sampling.

Even with the best sequencing approaches, PCR artifacts and sequencing
errors persist. This may account for some sequences not being observed
other samples. Although this lack of inter-sample coverage could have
been due to treatment effects and natural variation, it is likely that
some portion of the sequences that were unique to samples were artifacts
or contained errors. Researchers are encouraged to use rarefaction to
control for uneven sampling and the presence of spurious sequences.
Previous work sequencing mock communities has shown that the number of
spurious sequences increases with sequencing depth (2, 14). By rarefying
data to a common number of sequences per sample, the number of spurious
sequences can be controlled. As shown in the data I presented, which was
rarefied to a common number of reads per sample within a dataset, when
rare sequences were not removed the power to detect differences was the
highest and the false discovery rate was the expected 5\% (Figures 5 and
S6).

In addition to considerations of how to control for the presence of
spurious sequences, researchers also need to be mindful of how to
interpret the results of their work. Because every dataset will contain
residual sequences that are spurious, measures of richness and diversity
should be made on a relative basis. For example, pronouncements that
communities from an environment or treatment contain a specific number
of taxa are problematic. Instead, we should limit ourselves to
indicating that samples from one treatment group has more taxa than
another without using absolute values of richness. Furthermore, we must
take caution in interpreting rare taxa. Although the data from the
studies highlighted here suggest that most rare sequences are not
spurious, it is likely that some are. Therefore, researchers must
approach rare sequences with more skepticism than more abundant
sequences. Reseachers should seek out other methods to confirm
inferences that they make about rare sequences. This is a standard that
should be applied regardless of their abundance (15).

How to curate and interpret rare sequences has been a significant
challenge since microbial ecologists transitioned away from Sanger
sequencing of samples (16--18). Although the extent of the ``rare
biosphere'' is still an open question, it is important to appreciate the
importance of rare populations in all communities. Populations can be
numerically rare but ubuiquitous or abundant and limited in their
geographic range. Alternatively, they can be numerically rare but
temporally common or abundant but present infrequently. Removing
sequences from any of these settings will limit our ability to study the
role of such populations or the processes that drive their patchy
distributions that are so common with microbial communities (8).

@@

\textbf{Data curation and analysis.} To insure the highest possible data
quality, datasets were limited to those where the 500 cycle version 2
MiSeq chemistry was used to sequence the amplicons. The paired 250 nt
reads resulted in near complete 2-fold sequencing coverage of every
nucleotide in the ca. 253 nt-long region. This region and sequencing
platform were selected because previous work has shown that a standard
data analysis pipeline in mothur results in a sequencing error rate
below 0.02\% (2). All sequence data were obtained from the Sequence Read
Archive (SRA) and processed using a standard mothur-based sequencing
pipeline that resulted in ASVs as generated by the pre.cluster algorithm
using a threshold of 2 nt (2, 19). ASVs were assigned to OTUs using a
3\% distance threshold using mothur's cluster function with the
OptiClust algorithm (20). To minimize the effects of uneven sampling
effort, samples were rarefied to the number of sequences in the smallest
sample for each dataset. Because metrics of alpha diversity did not
consistently follow a normal distribution, I used the non-parametric
Wilcoxon rank test as implemented in R. For comparisons of Bray-Curtis
distances, the amova function within mothur was used, which implements
the analysis of molecular variance algorithm (21).

\textbf{Reproducibility.} All analyses were performed using mothur
(version 1.44.1) and R (version 4.0.2) with the tidyverse (version
1.3.0), broom (version 0.7.0), data.table (version 1.13.4), and cowplot
(version 1.1.0) packages. All of the code that was used in this analysis
as well as the Makefile for running the analysis are available at
\url{https://github.com/SchlossLab/Schloss_Singletons_XXXXX_2019} along
with the complete version history of the project.

@@

\leavevmode\hypertarget{ref-Reid2011}{}%
\CSLLeftMargin{8. }
\CSLRightInline{\textbf{Reid A}, \textbf{Buckley M}. 2011. The rare
biosphere.}

\leavevmode\hypertarget{ref-Morris2013}{}%
\CSLLeftMargin{9. }
\CSLRightInline{\textbf{Morris A}, \textbf{Beck JM}, \textbf{Schloss
PD}, \textbf{Campbell TB}, \textbf{Crothers K}, \textbf{Curtis JL},
\textbf{Flores SC}, \textbf{Fontenot AP}, \textbf{Ghedin E},
\textbf{Huang L}, \textbf{Jablonski K}, \textbf{Kleerup E},
\textbf{Lynch SV}, \textbf{Sodergren E}, \textbf{Twigg H}, \textbf{Young
VB}, \textbf{Bassis CM}, \textbf{Venkataraman A}, \textbf{Schmidt TM},
\textbf{Weinstock GM}. 2013. Comparison of the respiratory microbiome in
healthy nonsmokers and smokers. American Journal of Respiratory and
Critical Care Medicine \textbf{187}:1067--1075.
doi:\href{https://doi.org/10.1164/rccm.201210-1913oc}{10.1164/rccm.201210-1913oc}.}

\leavevmode\hypertarget{ref-Fadrosh2014}{}%
\CSLLeftMargin{10. }
\CSLRightInline{\textbf{Fadrosh DW}, \textbf{Ma B}, \textbf{Gajer P},
\textbf{Sengamalay N}, \textbf{Ott S}, \textbf{Brotman RM},
\textbf{Ravel J}. 2014. An improved dual-indexing approach for
multiplexed 16S {rRNA} gene sequencing on the illumina {MiSeq} platform.
Microbiome \textbf{2}:6.
doi:\href{https://doi.org/10.1186/2049-2618-2-6}{10.1186/2049-2618-2-6}.}

\leavevmode\hypertarget{ref-Holm2019}{}%
\CSLLeftMargin{11. }
\CSLRightInline{\textbf{Holm JB}, \textbf{Humphrys MS}, \textbf{Robinson
CK}, \textbf{Settles ML}, \textbf{Ott S}, \textbf{Fu L}, \textbf{Yang
H}, \textbf{Gajer P}, \textbf{He X}, \textbf{McComb E}, \textbf{Gravitt
PE}, \textbf{Ghanem KG}, \textbf{Brotman RM}, \textbf{Ravel J}. 2019.
Ultrahigh-throughput multiplexing and sequencing of \(>\)500-base-pair
amplicon regions on the {Illumina} {HiSeq} 2500 platform. {mSystems}
\textbf{4}.
doi:\href{https://doi.org/10.1128/msystems.00029-19}{10.1128/msystems.00029-19}.}
